$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-10 paper/paper.tex
Minor fix
Another pass to the introduction

  We present the syntax, semantics, type system and corresponding type safety properties for the \picalc{}.
  We generalise the algebras on multiplicities, and allow the user to use a mix of \emph{linear}, \emph{graded} and \emph{shared} types.
  We use leftover typing \cite{Allais2018a} to encode our typing rules, thus avoiding the ubiquitous context splitting proofs so common in type systems for linear calculi.
  We provide a framing theorem, and generalise the weakening and strengthening theorems, and use them to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda.
\end{abstract}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and (more recently) Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

J.-Y. Girard's development of linear logic \cite{Girard87} opened new avenues in computing science in the early '90s by introducing \emph{linear types} for (functional) programming languages \cite{Wadler90,Bernardy2018}.
Linear type systems guarantee that resources are used \emph{exactly once}.
Enforcing ``no duplicating'' and ``no discarding'' of resources via linearity allows for resource-aware programming and more efficient implementations \cite{Wadler90}.
Linearity later on inspired unique types (as in Clean \cite{BarendsenS96}) and ownership types (as in Rust \cite{MatsakisK14}).

Kobayashi et al. \cite{KPT96} brought linearity to the \picalc{} and provided a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
In the meantime, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98} --- a formalism used in communication-centric programming.
With session types, linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself is used multiple times, as per its session type.
Linearity in session types is the key ingredient that guarantees type safety.

Recently, an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} into the spotlight again \cite{KPT96}.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but is useful in practice too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}, and it also allows us to use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be built.

@@

We do so by defining a specification for an algebra on \emph{multiplicities} (how many times a channel can be used) and use it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be used either separately or mixed in a type system.
Ultimately, we can exploit this formalisation as a unified framework for type safe distributed modelling and programming with a variety of type systems, and build on top of it more advanced types, theories and languages.

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level \emph{de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it.
     For convenience, the user can convert from and to a syntax with names.
     These conversion functions are shown to be inverses of one another modulo $\alpha$-conversion.

   \item \textbf{Semantics}: \autoref{semantics} provides an operational semantics for the \picalc{}, prior to any typing.
   This operational semantics is defined as a reduction relation on processes (\autoref{operational-semantics}).
   The reduction relation tracks at the type level the channel the communication occurs on, so that this information can then be used to state subject reduction --- aka type preservation (\autoref{thm:subject-reduction}).
   The reduction relation is defined modulo structural congruence (\autoref{structural-congruence}) --- a relation that acts as a quotient type that removes unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

  \item \textbf{Leftover typing for the \picalc{} with linear, graded and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
    Linear types, graded types and shared types are all instances of the same resource-aware algebra, giving rise to three different type systems under the same framework.
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra.

@@

    \item \textbf{\picalc{} with leftovers}: Our type system uses \emph{leftover typing} to model the resource-aware \picalc{} (\autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgements.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.

@@

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{Four communicating processes composed in parallel.}
  \label{fig:example-process}
\end{figure}

The example contains four processes that are composed in parallel: two of them send data over a channel; one receives data twice; the fourth sticks together and synchronizes the previous three processes (\autoref{fig:example-process}).
The process doing the synchronisation first receives on both input channels and then sends the received data along the output channel.
Thanks to linear and graded types, the processes below can have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{sync}$ consumes one input multiplicity on the channels shared with $\func{send}$, and two output multiplicities on the channel shared with $\func{recv}$.
We use these types to give a suitable typing derivation in \autoref{example-derivation}.

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
Constructors ($\constr{0}$ and $\suc$) are given as inference rules, although we occasionally use a BNF style definition of types for brevity.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda --- where premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned in type definitions but omitted by constructors.

We use colours to further distinguish the different entities in this paper.
$\type{TYPES}$ are blue (uppercased, with indices as subscripts), $\constr{constructors}$ are orange, $\func{functions}$ are olive green, variables are black, and some constructor names are overloaded --- and disambiguated by context.

@@

    \inferrule
    {eq : P \eq{r} P'}
    {\constr{cong-recv} \; eq : \recv{x}P \eq{\constr{one} \; r} \recv{x}P'}

>>>>

  We present an untyped syntax and semantics for the \picalc{}.
  We provide a resource-aware type system parametrised by a set of multiplicity algebras.
  This allows the user to mix \emph{linear}, \emph{graded} and \emph{shared} types.
  We use leftover typing \cite{Allais2018a} to encode our typing rules, thus avoiding the ubiquitous context splitting proofs so common in type systems for linear calculi.
  We provide a framing theorem, and generalise the weakening and strengthening theorems, and use them to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda.
\end{abstract}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be transmitted.
The \picalc{} is a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and (more recently) Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types.
These make the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.

In the early '90s, \emph{linear types} for (functional) programming languages \cite{Wadler90,Bernardy2018} were introduced, following J.-Y. Girard's development of linear logic \cite{Girard87}.
Linear type systems guarantee that resources are used \emph{exactly once}.
Enforcing the ``no duplicating'' and ``no discarding'' of resources via linearity allows for resource-aware programming and more efficient implementations \cite{Wadler90}.
Later on, linearity inspired unique types (as in Clean \cite{BarendsenS96}) and ownership types (as in Rust \cite{MatsakisK14}).

Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98} (a formalism used in communication-centric programming), where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself is used multiple times, as per its session type.
Linearity in session types is the key ingredient that guarantees type safety.

Recently, an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} into the spotlight again \cite{KPT96}.
This encoding not only has theoretical benefits in terms of the expressivity of session types and the reusability of theoretical results from the linear \picalc{}, but is useful in practice too.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}, and it allows to use the linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be built.

@@

We do so by defining a specification for an algebra on \emph{multiplicities} (how many times a channel can be used) and use it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be mixed in the same type system.
Ultimately, we can exploit this formalisation as a unified framework for type-safe distributed modelling and programming using a variety of type systems, and build on top of it more advanced types, theories and languages.

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level \emph{de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it.
     For convenience, the user can convert from and to a syntax with names.
     These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.

   \item \textbf{Semantics}: \autoref{semantics} provides an operational semantics for the \picalc{}, prior to any typing.
   This operational semantics is defined as a reduction relation on processes (\autoref{operational-semantics}).
   The reduction relation tracks at the type level the channel the communication occurs on, so that this information can later be used to state subject reduction --- aka type preservation (\autoref{thm:subject-reduction}).
   The reduction relation is defined modulo structural congruence (\autoref{structural-congruence}) --- a relation that acts as a quotient type that removes unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

  \item \textbf{Leftover typing for the \picalc{} with support for linear, graded and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra --- linear types, graded types and shared types all satisfy these algebraic laws.

@@

    \item \textbf{Leftover typing}: Our type system uses \emph{leftover typing} to model the resource-aware \picalc{} (\autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgements.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the remaining as leftovers in the output usage context.

@@

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{Four communicating example processes composed in parallel.}
  \label{fig:example-process}
\end{figure}

The example contains four processes that are composed in parallel: two of them send data over a channel; one receives data twice; the fourth sticks together and synchronizes the previous three processes (\autoref{fig:example-process}).
The process doing the synchronisation first receives on both input channels and then sends the received data along the output channel.
Thanks to linear and graded types, the processes below can have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{sync}$ consumes one input multiplicity on the channels shared with $\func{send}$, and two output multiplicities on the channel shared with $\func{recv}$.
We use these types to give a suitable typing derivation in \autoref{example-derivation} (\autoref{leftover-typing}).

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double inference lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
Constructors ($\constr{0}$ and $\suc$) are given as inference rules.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda: premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned in type definitions but omitted by constructors.

We use colours to further distinguish the different entities in this paper.
$\type{TYPES}$ are blue and uppercased, with indices as subscripts, $\constr{constructors}$ are orange, $\func{functions}$ are olive green, variables are black, and some constructor names are overloaded --- and disambiguated by context.

@@

    \inferrule
    {eq : P \eq{r} P'}
    {\constr{cong-recv} \; eq : \recv{x}{}P \eq{\constr{one} \; r} \recv{x}{}P'}
